\documentclass{article}
\usepackage{natbib}
\usepackage{todonotes}

\title{\bf Final Project Proposal}
\author{Your Name}
\date{Due date: 8 December 2015 at 11:59 PM Central time}

\begin{document}
\maketitle


% Lines that begin with % are comments.


\section{Motivation}
\label{sec:motivation}

% Delete this todo
\todo[inline]{
Write an introductory section that motivates a particular problem you wish to solve. This section should (at a minimum) fill the remainder of this page.
}


\section{Related Work}

% Delete this todo
\todo[inline]{%
Briefly survey the most salient prior work that relates to the problem you wish to solve. This section should cite relevant sources. This section should be at half a page in length.
}


\section{Proposed Work}

% Delete this todo
\todo[inline]{This section should contain a detailed description of the work you propose to do, but it should not contain any experimental results. This section should take at least a page and a half.}



\section{Results}

% Delete this todo
\todo[inline]{This section should describe the your results. In other words, after you do what you proposed in section 3, what results did you come up with? This section should be at least one page in length.
}


\section{Conclusion}

% Delete this todo
\todo[inline]{Write a concluding summary. Also indicate what future work may be warranted. This section should be at least half a page in length.}


\bibliographystyle{apalike}
\bibliography{your_bibliography}

\end{document}